---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
set.seed(123)
```

# `xplainfi`

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/jemus42/xplainfi/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jemus42/xplainfi/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of `xplainfi` is to collect common feature importance methods under a unified and extensible interface.  

For now, it is built specifically around [mlr3](https://mlr-org.com/), as available abstractions for learners, tasks, measures, etc. greatly simplify the implementation of importance measures.

## Installation

You can install the development version of `xplainfi` like so:

``` r
# install.packages(pak)
pak::pak("jemus42/xplainfi")
```

## Example

Here is a basic example on how to calculate PFI for a given learner and task:

```{r example}
library(xplainfi)
library(mlr3)
library(mlr3learners)

task = tsk("german_credit")
learner = lrn("classif.ranger", num.trees = 500)
measure = msr("classif.ce")

pfi = PFI$new(
  task = task,
  learner = learner,
  measure = measure,
  resampling = rsmp("cv", folds = 3)
)
pfi$compute()
```

