---
title: "Perturbation-based Feature Importance Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Perturbation-based Feature Importance Methods}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(123)
# Quiet down
lgr::get_logger("mlr3")$set_threshold("warn")
```

```{r setup}
library(xplainfi)

# learners, tasks, etc.
library(mlr3)
library(mlr3data) # penguins_simple
library(mlr3learners) # ranger
library(data.table)
```

This vignette demonstrates the three perturbation-based feature importance methods implemented in xplainfi:

- **PFI (Permutation Feature Importance)**: Uses marginal sampling (simple permutation)
- **CFI (Conditional Feature Importance)**: Uses conditional sampling via Adversarial Random Forests
- **RFI (Relative Feature Importance)**: Uses conditional sampling on a user-specified subset of features

## Problem Setup

We'll use the `penguins_simple` dataset to compare these methods:

```{r setup-problem}
task <- tsk("penguins")
learner <- lrn("classif.ranger", num.trees = 100, num.threads = 2)
resampling <- rsmp("cv", folds = 3)
measure <- msr("classif.acc")

# Show the task structure
task
```

## Permutation Feature Importance (PFI)

PFI shuffles each feature independently, breaking the association between the feature and the target while preserving the marginal distribution.

For more stable results, we 

- a) Use 3-fold CV for more reliable performance estimates
- b) Within each resampling iteration, we repeat the permutation-prediction-scoring step `iters_perm` times


```{r pfi}
pfi <- PFI$new(
  task = task,
  learner = learner,
  measure = measure,
  resampling = resampling,
  iters_perm = 5
)

# Compute importance scores
pfi_results <- pfi$compute(relation = "difference")
pfi_results

# Also stored in
pfi$importance

# Scores per resampling iteration and permutation iteration
pfi$scores |>
  knitr::kable(digits = 4)
```

## Conditional Feature Importance (CFI)

CFI uses conditional sampling to preserve the joint distribution of all other features when perturbing a feature of interest. By default we use Adversarial Random Forests (ARF) as conditional sampler internally, but it can be provided explicity, as we usually want to supply additional arguments to `arf::adversarial_rf()` or `arf::forde()`.

```{r sampler-arf}
sampler = ARFSampler$new(
  task = task, 
  arf_args = list(verbose = FALSE, parallel = FALSE),
  forde_args = list(finite_bounds = "no")
)

# Example sampling for 5 randomly chosen rows from the task
sampler$sample(
  feature = "bill_length", 
  data = task$data(rows = sample(task$nrow, size = 5)), 
  conditioning_features = "body_mass"
)
```

```{r cfi}
cfi <- CFI$new(
  task = task,
  learner = learner,
  measure = measure,
  resampling = resampling,
  iters_perm = 5,
  sampler = sampler
)

# Compute importance scores
cfi_results <- cfi$compute(relation = "difference")
cfi_results
```

## Relative Feature Importance (RFI)

RFI conditions on a specific subset of features as opposed to just one as in CFI, measuring importance relative to those features.

```{r rfi}
conditioning_set <- c("sex", "year")

rfi <- RFI$new(
  task = task,
  learner = learner,
  measure = measure,
  resampling = resampling,
  conditioning_set = conditioning_set,
  iters_perm = 5,
  sampler = sampler
)

# Compute importance scores
rfi_results <- rfi$compute(relation = "difference")
rfi_results
```

## Comparing Methods

Now let's compare the results from all three methods:

```{r comparison}
# Combine results for comparison
comparison <- merge(
  pfi_results[, .(feature, pfi = importance)],
  cfi_results[, .(feature, cfi = importance)],
  by = "feature"
)
comparison <- merge(
  comparison,
  rfi_results[, .(feature, rfi = importance)],
  by = "feature"
)

comparison
```

```{r comparison-table}
comparison |>
  knitr::kable(
    digits = 4, 
    caption = "Feature Importance Comparison (Difference Scores)",
    col.names = c("Feature", "PFI", "CFI", "RFI")
  )
```

## Visualization

```{r visualization, fig.width=8, fig.height=6}
library(ggplot2)

# Reshape data for plotting
plot_data <- comparison |>
  melt(
    id.vars = "feature",
    value.name = "importance",
    variable.name = "method"
  )

# Clean up method names
plot_data[, method := toupper(method)]

# Create the plot
ggplot(plot_data, aes(x = importance, y = reorder(feature, importance), fill = method)) +
  geom_col(position = "dodge", alpha = 0.7) +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "Feature Importance Comparison",
    subtitle = glue::glue(
      "Task: {task$id} 
      Measure: {measure$id}
      RFI conditioned on: {paste(conditioning_set, collapse = ', ')}"),
    x = "Importance Score (Difference)",
    y = "Feature",
    fill = "Method",
    caption = glue::glue("Using {resampling$iters}-fold cross-validation")
  ) +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )
```

## Understanding the Differences

The three methods can yield different results because they make different assumptions:

1. **PFI** assumes features are independent and measures the marginal importance of each feature.

2. **CFI** preserves the conditional distribution $P(X_{-j} | X_j)$ when perturbing feature $j$, providing a more realistic assessment when features are correlated.

3. **RFI** measures importance relative to a specific conditioning set, answering "what additional information does this feature provide beyond what we already know from features A, B, C?"

In our penguin example:
- Features like bill dimensions and body mass are likely correlated
- CFI accounts for these correlations when measuring importance
- RFI shows which features add value beyond bill_length and flipper_length

## Advanced Usage

### Custom Samplers

You can also provide custom samplers for more control:

```{r custom-sampler, eval=FALSE}
# Create a custom ARF sampler with specific parameters
custom_sampler <- ARFSampler$new(
  task = task,
  arf_args = list(num_trees = 50),  # Fewer trees for faster computation
  forde_args = list(finite_bounds = "local") # Avoid out of distribution samples
)

cfi_custom <- CFI$new(
  task = task,
  learner = learner,
  measure = measure,
  resampling = resampling,
  sampler = custom_sampler
)
```

### Accessing Detailed Results

As mentioned, individual scores per resampling iteration and permutation iteration are kept in `$scores`.

```{r detailed-scores}
detailed_scores <- pfi$scores
head(detailed_scores)

# Summarize variability across iterations
score_summary <- detailed_scores[, .(
  mean_importance = mean(importance),
  sd_importance = sd(importance),
  n_iterations = .N
), by = feature]

score_summary

ggplot(score_summary, aes(
  y = reorder(feature, mean_importance), 
  xmin = mean_importance - sd_importance, 
  xmax = mean_importance + sd_importance)
  ) +
  geom_errorbarh(linewidth = 1, height = .5) +
  labs(
    title = "Permutation Feature Importance: Mean ± SD",
    subtitle = glue::glue(
      "Task: {task$id} 
      Measure: {measure$id}"),
    x = "PFI Score (Difference)",
    y = "Feature",
    caption = glue::glue("Using {resampling$iters}-fold cross-validation and {pfi$param_set$values$iters_perm} permutations each")
  ) +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )
```
