---
title: "xplainvi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{xplainvi}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xplainvi)

# learners, tasks, etc.
library(mlr3)
library(mlr3learners)
```

Defining a simple example case:

- German credit classification task
- Random forests with 100 trees
- Holdout split (TBI)
- Measure: Classification error

```{r setup-problem}
task = tsk("german_credit")
learner = lrn("classif.ranger", num.trees = 500)
measure = msr("classif.ce")
```

## Simple case without resampling

Default behavior will internally construct standard holdout resampling with default ratio

Calculating PFI: 

```{r}
pfi = PFI$new(
  task = task, 
  learner = learner,
  measure = measure
)

# Stores parameter set to calculate PFI in different ways
pfi$param_set

set.seed(123)

# Default behavior should be sane
pfi$compute()
```

Does not recompute if not needed

```{r}
pfi$compute(relation = "difference")
```

Recomputes if param changes, stores new param

```{r}
pfi$compute(relation = "ratio")
pfi$param_set
```

Caveat: When `$compute()` is called again its default value for `"relation"` (i.e. `"difference"`) is used

```{r}
pfi$compute()
```

Retrieve scores and convert to DT:

```{r}
pfi$importance
as.data.table(pfi)
```

## With resampling

```{r}
learner = lrn("classif.ranger", num.trees = 100)
resampling = rsmp("cv", folds = 3)
measure = msr("classif.ce")


pfi = PFI$new(
  task = task, 
  learner = learner, 
  resampling = resampling,
  measure = measure
)

pfi$resampling
pfi$resample_result

pfi$compute(relation = "difference")

pfi$resample_result

pfi$importance

```

Different measure:  

(Note: Maybe it would be worth allowing to change measure post-hoc?)

```{r}
learner$predict_type = "prob"

pfi = PFI$new(
  task = task, 
  learner = learner, 
  resampling = resampling,
  measure = msr("classif.auc")
)

pfi$compute(relation = "ratio")
pfi$compute(relation = "difference")

```

