---
title: "Comparison with fippy (Python Implementation)"
subtitle: "Cross-language validation of feature importance methods"
editor_options: 
  chunk_output_type: console
---

```{r setup, message=FALSE}
library(data.table)
library(ggplot2)
library(jsonlite)
library(knitr)
library(glue)
library(here)

# Set seed for reproducibility
set.seed(123)
```

## Overview

This article compares xplainfi's feature importance implementations with those from [fippy](https://github.com/gcskoenig/fippy), a Python package implementing similar methods. This comparison serves as a regression test to ensure methodological consistency across language implementations.

The comparison includes:

- **PFI** (Permutation Feature Importance)
- **CFI** (Conditional Feature Importance) 
- **RFI** (Relative Feature Importance)
- **SAGE** (Shapley Additive Global Importance) - both marginal and conditional variants

## Methodology

Both implementations use:

- **Dataset**: Friedman1 task with 1000 observations (`mlbench::mlbench.friedman1()` via `mlr3::tgen("friedman1")`)
- **Model**: Random Forest with 100 trees
- **Evaluation**: Same train/test data
- **Metrics**: Mean Squared Error for importance calculations

Due to the difference in underlying conditional samplers (ARF for xplainfi) we expect conditional methods to be less similar than marginal ones.

## Setup and Execution

The comparison uses separate calculation scripts:

```{bash}
#| eval: false

# 1. Calculate xplainfi results
cd vignettes/articles/fippy-comparison
Rscript calculate_xplainfi.R

# 2. Calculate fippy results (portable with uv - automatically installs dependencies)
./calculate_fippy.py
```

Both scripts generate JSON files with results that are loaded below for comparison.

## Load Results

```{r load-results}
# Check that both result files exist
# Look in the fippy-comparison subdirectory
base_dir <- here::here("vignettes", "articles", "fippy-comparison")
xplainfi_results_path <- file.path(base_dir, "xplainfi_results.json")
fippy_results_path <- file.path(base_dir, "fippy_results.json")

if (!file.exists(xplainfi_results_path)) {
  stop("xplainfi_results.json not found. Please run calculate_xplainfi.R first.")
}

if (!file.exists(fippy_results_path)) {
  stop("fippy_results.json not found. Please run calculate_fippy.py first.")
}

# Load results from both implementations
xplainfi_results <- fromJSON(xplainfi_results_path)
fippy_results <- fromJSON(fippy_results_path)
```

## Model Performance Comparison

```{r model-performance}
performance_comparison <- data.table(
  Implementation = c("xplainfi (R)", "fippy (Python)"),
  R_squared = c(
    round(xplainfi_results$model_performance$r_squared, 4),
    round(fippy_results$model_performance$r_squared, 4)
  )
)

kable(performance_comparison, caption = "Model Performance Comparison")
```

## Method Comparisons

```{r comparison-function}
compare_method <- function(method_name, xplainfi_result, fippy_result) {
  # Both implementations available
  method_dt <- data.table(
    feature = xplainfi_result$feature,
    xplainfi = xplainfi_result$importance,
    fippy = fippy_result$importance
  )
  
  # Return table and correlation for display
  correlation <- cor(method_dt$xplainfi, method_dt$fippy)
  correlation_spearman = cor(method_dt$xplainfi, method_dt$fippy, method = "spearman")
  
  list(
    method = method_name, 
    table = kable(method_dt[order(-xplainfi)], 
                  caption = glue("{method_name} Results Comparison"), 
                  digits = 4),
    correlation = correlation,
    correlation_spearman = correlation_spearman
  )
}
```

### PFI (Permutation Feature Importance)

```{r pfi-comparison}
pfi_result <- compare_method("PFI", xplainfi_results$PFI, fippy_results$PFI)
pfi_result$table

glue("PFI Correlation: {round(pfi_result$correlation, 3)}")

pfi_result$correlation
pfi_result$correlation_spearman
```

### CFI (Conditional Feature Importance)

```{r cfi-comparison}
cfi_result <- compare_method("CFI", xplainfi_results$CFI, fippy_results$CFI)
cfi_result$table

glue("CFI Correlation: {round(cfi_result$correlation, 3)}")

cfi_result$correlation
cfi_result$correlation_spearman
```

### RFI (Relative Feature Importance)

```{r rfi-comparison}
rfi_result <- compare_method("RFI", xplainfi_results$RFI, fippy_results$RFI)
rfi_result$table

glue("RFI Correlation: {round(rfi_result$correlation, 3)}")

rfi_result$correlation
rfi_result$correlation_spearman

glue("xplainfi conditioning set: {paste(xplainfi_results$RFI$conditioning_set, collapse = ', ')}")
glue("fippy conditioning set: {paste(fippy_results$RFI$conditioning_set, collapse = ', ')}")
```

### SAGE Methods

#### Marginal SAGE

```{r sage-marginal-comparison}
sage_marginal_result <- compare_method("Marginal SAGE", xplainfi_results$SAGE_Marginal, fippy_results$SAGE_Marginal)
sage_marginal_result$table

glue("Marginal SAGE Correlation: {round(sage_marginal_result$correlation, 3)}")

sage_marginal_result$correlation
sage_marginal_result$correlation_spearman
```

#### Conditional SAGE

```{r sage-conditional-comparison}
sage_conditional_result <- compare_method("Conditional SAGE", xplainfi_results$SAGE_Conditional, fippy_results$SAGE_Conditional)
sage_conditional_result$table

glue("Conditional SAGE Correlation: {round(sage_conditional_result$correlation, 3)}")

sage_conditional_result$correlation
sage_conditional_result$correlation_spearman
```

## Correlation Summary

```{r correlation-summary}
correlations <- rbindlist(list(
  pfi_result[c("method", "correlation", "correlation_spearman")],
  cfi_result[c("method", "correlation", "correlation_spearman")],
  rfi_result[c("method", "correlation", "correlation_spearman")],
  sage_marginal_result[c("method", "correlation", "correlation_spearman")],
  sage_conditional_result[c("method", "correlation", "correlation_spearman")]
))

kable(
  correlations, 
  caption = "Pearson and Spearman Correlations between xplainfi and fippy", 
  col.names = c("Method", "Pearson Corr.", "Spearman Corr.")
)

melt(correlations, id.vars = "method") |>
  ggplot(aes(x = reorder(method, value), y = value)) +
    facet_wrap(vars(variable), ncol = 1, labeller = as_labeller(c(correlation = "Pearson's", correlation_spearman = "Spearman's"))) +
    geom_col(fill = "steelblue", alpha = 0.7) +
    geom_hline(yintercept = c(0.5, 1), linetype = "dashed", color = "red", alpha = 0.7) +
    coord_flip() +
    labs(
      title = "Implementation Correlations",
      subtitle = "xplainfi (R) vs fippy (Python)",
      x = "Method",
      y = "Correlation"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title.position = "plot"
    )

```
